---
title: "Introduction to dsHPC"
author: "David Sarrat González, Xavier Escribà-Montagut, Juan R González"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Introduction to dsHPC}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

## Introduction

The `dsHPC` package provides a seamless interface between DataSHIELD server-side operations and High-Performance Computing (HPC) clusters. It enables computationally intensive analyses to be offloaded to HPC resources, with caching mechanisms to avoid redundant calculations.

This vignette introduces the basic usage of the package for both end-users and developers creating other DataSHIELD packages.

## Basic Workflow

The typical workflow for using `dsHPC` consists of:

1. Initializing the system
2. Submitting jobs
3. Checking job status
4. Retrieving results

### Initialization

Before using any other functions in the package, you need to initialize it:

```{r eval=FALSE}
library(dsHPC)
dsHPC.init()
```

This sets up the database connection for caching and checks for available job schedulers. If no Slurm cluster is available, the package will fallback to local execution.

### Submitting Jobs

You can submit a function to be executed on the HPC cluster:

```{r eval=FALSE}
# Submit a simple job
job <- dsHPC.submit(
  func_name = "mean",
  args = list(x = c(1, 2, 3, 4, 5))
)

# The job object contains information about the submission
job
```

For more complex operations, you can specify additional parameters:

```{r eval=FALSE}
# Submit a job with more options
job <- dsHPC.submit(
  func_name = "kmeans",
  args = list(x = my_data_matrix, centers = 3),
  package = "stats",
  required_packages = c("stats"),
  use_cache = TRUE,
  slurm_opts = list(
    partition = "normal",
    memory = "8G",
    time = "02:00:00",
    cpus = 2
  )
)
```

### Checking Job Status

You can check the status of a submitted job:

```{r eval=FALSE}
# Get job status
status <- dsHPC.status(job$job_id)
status
```

The status will be one of:
- `SUBMITTED`: Job has been submitted to the scheduler
- `PENDING`: Job is waiting to be executed
- `RUNNING`: Job is currently running
- `COMPLETED`: Job has completed successfully
- `FAILED`: Job has failed
- `CACHED`: Result retrieved from cache

### Retrieving Results

Once a job is completed, you can retrieve the results:

```{r eval=FALSE}
# Get job result (non-blocking)
result <- dsHPC.result(job$job_id)

# Get job result (blocking, waits for completion)
result <- dsHPC.result(job$job_id, wait_for_completion = TRUE)
```

### Job Management

You can list all jobs or filter by status:

```{r eval=FALSE}
# List all jobs
all_jobs <- dsHPC.list_jobs()

# List only completed jobs
completed_jobs <- dsHPC.list_jobs(status_filter = "COMPLETED")
```

Cancel a running job:

```{r eval=FALSE}
# Cancel job
dsHPC.cancel(job$job_id)
```

Clean up old cached results:

```{r eval=FALSE}
# Remove cache entries older than 14 days
dsHPC.clean_cache(days_to_keep = 14)
```

## DataSHIELD Integration

For DataSHIELD environments, use the DataSHIELD-specific functions:

```{r eval=FALSE}
# Initialize in DataSHIELD environment
dsHPC.ds.init()

# Submit a job in DataSHIELD environment
job <- dsHPC.ds.submit(
  func_name = "kmeans",
  args = list(x = my_data_matrix, centers = 3),
  package = "stats"
)

# Get job status in DataSHIELD environment
status <- dsHPC.ds.status(job$job_id)

# Get job result in DataSHIELD environment
result <- dsHPC.ds.result(job$job_id, wait_for_completion = TRUE)
```

## Function Wrapping for Easier Use

You can create a wrapped version of a function that will automatically be executed via HPC:

```{r eval=FALSE}
# Create a wrapped version of the kmeans function
kmeans_hpc <- dsHPC.wrap_function(
  func = kmeans,
  package = "stats",
  required_packages = c("stats"),
  slurm_opts = list(memory = "8G", cpus = 2)
)

# Use the wrapped function as you would use the original function
result <- kmeans_hpc(x = my_data, centers = 3)
```

This is particularly useful when you want to transparently offload computation-intensive functions to HPC resources.

## For Package Developers

If you're developing a DataSHIELD server-side package and want to leverage HPC capabilities, you can integrate `dsHPC` as follows:

1. Add `dsHPC` as a dependency in your package
2. Initialize `dsHPC` if needed
3. Use `dsHPC.submit()` or `dsHPC.wrap_function()` to execute computation-intensive functions

Example:

```{r eval=FALSE}
#' My intensive analysis function
#' 
#' @param data Data frame to analyze
#' @param ... Additional parameters
#' @export
my_intensive_analysis <- function(data, ...) {
  # Initialize dsHPC if not already done
  if (is.null(getOption("dsHPC.config"))) {
    dsHPC::dsHPC.init()
  }
  
  # Submit the actual analysis function to HPC
  job <- dsHPC::dsHPC.submit(
    func_name = "actual_analysis_implementation",
    args = list(data = data, ...),
    package = "myPackage"
  )
  
  # Wait for and return the result
  dsHPC::dsHPC.result(job$job_id, wait_for_completion = TRUE)
}
```

This approach allows your package to benefit from HPC resources while maintaining a simple API for end users.

## Conclusion

The `dsHPC` package provides a powerful and flexible interface for offloading computational tasks to HPC resources within the DataSHIELD framework. By handling job submission, monitoring, and result caching, it enables efficient use of computing resources while maintaining ease of use.

For more detailed information about specific functions, please refer to the package documentation. 