---
title: "Job Management with dsHPC"
author: "David Sarrat González, Xavier Escribà-Montagut, Juan R González"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Job Management with dsHPC}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval = FALSE
)
```

## Introduction

The `dsHPC` package provides comprehensive job management capabilities to help you track and interact with computational jobs throughout their lifecycle. This vignette demonstrates how to monitor, query, and manage both R and Python jobs across various states.

## Job States

Jobs in dsHPC can exist in one of the following states:

1. **SUBMITTED**: The job has been submitted to the scheduler or processing queue
2. **PENDING**: The job is waiting in the scheduler queue
3. **RUNNING**: The job is currently executing
4. **COMPLETED**: The job has finished successfully
5. **FAILED**: The job encountered an error during execution
6. **CANCELLED**: The job was manually cancelled
7. **CACHED**: The result was retrieved from cache (not a real job state, but reported by the status function)

## Basic Setup

Before managing jobs, you need to initialize the dsHPC environment:

```{r setup}
library(dsHPC)
dsHPC.init()
```

## Submitting Jobs

### R Jobs

```{r submit_r_job}
# Define a function that takes some time to run
long_computation <- function(size) {
  # Simulate a long computation
  result <- matrix(rnorm(size^2), size, size)
  # Compute eigenvectors (computationally intensive for large matrices)
  eigen_result <- eigen(result)
  return(eigen_result$values)
}

# Submit the job
r_job <- dsHPC.submit(
  func = long_computation,
  args = list(size = 1000),
  slurm_opts = list(
    memory = "4G",
    cpus = 2
  )
)

# The job is now submitted and has a unique ID
print(r_job$job_id)
```

### Python Jobs

```{r submit_python_job}
# For Python jobs, we use dsHPC.submit_python
py_job <- dsHPC.submit_python(
  py_module = "sklearn.cluster",
  py_function = "KMeans",
  args = list(
    n_clusters = 5,
    random_state = 42
  ),
  required_modules = c("numpy", "sklearn")
)

# Python jobs also receive a unique ID
print(py_job$job_id)
```

## Checking Job Status

The most basic task in job management is checking the current status of a job:

```{r check_status}
# Check status of R job
r_status <- dsHPC.status(r_job$job_id)
print(r_status)

# Check status of Python job
py_status <- dsHPC.status(py_job$job_id)
print(py_status)
```

The `dsHPC.status()` function returns a list containing information about the job, including:

- `job_id`: The unique identifier of the job
- `job_hash`: A hash based on the function and arguments
- `status`: The current state of the job ("SUBMITTED", "PENDING", "RUNNING", "COMPLETED", "FAILED", or "CANCELLED")
- `result`: The job result, if available and requested (see below)

## Retrieving Results

Once a job is completed, you can retrieve its result:

```{r get_result}
# Basic result retrieval (non-blocking)
if (r_status$status == "COMPLETED") {
  result <- dsHPC.result(r_job$job_id)
  print(head(result))  # Show the first few eigenvalues
}
```

### Waiting for Completion

You can also wait for a job to complete and then retrieve the result in a single call:

```{r wait_for_result}
# Waiting for completion (blocking)
result <- dsHPC.result(r_job$job_id, wait_for_completion = TRUE)

# You can also specify a timeout in seconds
result <- dsHPC.result(py_job$job_id, 
                       wait_for_completion = TRUE, 
                       timeout = 300)  # 5-minute timeout
```

The `wait_for_completion` parameter causes the function to block until the job completes or the timeout is reached. If the timeout is reached, the function returns `NULL` and issues a warning.

### Checking Completion Status and Result in One Call

You can retrieve both the job status and its result (if available) in a single call:

```{r status_with_result}
# Check status and get result if available
status_with_result <- dsHPC.status(r_job$job_id, return_result = TRUE)

if (status_with_result$status == "COMPLETED") {
  print(head(status_with_result$result))
}
```

## Listing Jobs

You can list all jobs or filter by status:

```{r list_jobs}
# List all jobs
all_jobs <- dsHPC.list_jobs()
print(all_jobs)

# List only completed jobs
completed_jobs <- dsHPC.list_jobs(status_filter = "COMPLETED")
print(completed_jobs)

# List jobs that are either running or pending
active_jobs <- dsHPC.list_jobs(status_filter = c("RUNNING", "PENDING"))
print(active_jobs)
```

The `dsHPC.list_jobs()` function returns a data frame containing information about all jobs that match the filter criteria.

## Cancelling Jobs

If you want to stop a job that's currently running or pending:

```{r cancel_job}
# Cancel a running job
if (r_status$status %in% c("RUNNING", "PENDING")) {
  cancelled <- dsHPC.cancel(r_job$job_id)
  if (cancelled) {
    print("Job cancelled successfully")
  } else {
    print("Failed to cancel job")
  }
}
```

Note that job cancellation only works if the job is in a cancellable state (SUBMITTED, PENDING, or RUNNING) and if the job scheduler is available. In simulation mode, or for jobs that have already completed or failed, cancellation will not work.

## Handling Failed Jobs

When a job fails, you can still retrieve information about the failure:

```{r failed_job}
# Check if a job failed
if (py_status$status == "FAILED") {
  # Get detailed job information
  py_job_info <- get_job_info(getOption("dsHPC.config")$connection, py_job$job_id)
  
  # Print the error message
  print(paste("Job failed with error:", py_job_info$error_message))
}
```

## Practical Workflow for Job Management

Here's a practical workflow for managing multiple jobs:

```{r practical_workflow}
# Submit multiple jobs
jobs <- list()
for (i in 1:5) {
  jobs[[i]] <- dsHPC.submit(
    func = function(n) sum(rnorm(n)^2),
    args = list(n = i * 1000)
  )
}

# Monitor all jobs until completion
monitor_until_completion <- function(job_list, interval = 2, timeout = 300) {
  start_time <- Sys.time()
  while (TRUE) {
    # Check all jobs
    all_completed <- TRUE
    for (i in seq_along(job_list)) {
      status <- dsHPC.status(job_list[[i]]$job_id)
      cat(sprintf("Job %d: %s\n", i, status$status))
      
      if (!status$status %in% c("COMPLETED", "FAILED", "CANCELLED", "CACHED")) {
        all_completed <- FALSE
      }
    }
    
    # If all jobs are in a final state, exit the loop
    if (all_completed) {
      break
    }
    
    # Check timeout
    elapsed <- difftime(Sys.time(), start_time, units = "secs")
    if (elapsed > timeout) {
      warning("Timeout reached while waiting for job completion")
      break
    }
    
    # Wait before checking again
    Sys.sleep(interval)
  }
  
  # Collect all results
  results <- list()
  for (i in seq_along(job_list)) {
    status <- dsHPC.status(job_list[[i]]$job_id)
    if (status$status == "COMPLETED" || status$status == "CACHED") {
      results[[i]] <- dsHPC.result(job_list[[i]]$job_id)
    } else {
      results[[i]] <- NULL
    }
  }
  
  return(results)
}

# Monitor all jobs and collect results
results <- monitor_until_completion(jobs)
print(results)
```

## Implementing Job Management in DataSHIELD Packages

If you're developing a DataSHIELD package that uses dsHPC, consider implementing these job management patterns:

```{r datashield_package}
# Example of a function for a DataSHIELD package
myPackage.submitAnalysis <- function(data, params) {
  # Check if dsHPC is available
  if (!requireNamespace("dsHPC", quietly = TRUE) || 
      is.null(getOption("dsHPC.config"))) {
    stop("dsHPC is not available or initialized")
  }
  
  # Submit the job
  job <- dsHPC.submit(
    func = function(data, params) {
      # Actual computation code
      result <- analyze_data(data, params)
      return(result)
    },
    args = list(data = data, params = params),
    use_cache = TRUE
  )
  
  # Return job ID and info for status checking
  return(list(
    job_id = job$job_id,
    status = job$status,
    message = "Analysis job submitted successfully. Use myPackage.checkAnalysis() to check status."
  ))
}

# Function to check analysis status
myPackage.checkAnalysis <- function(job_id) {
  if (!requireNamespace("dsHPC", quietly = TRUE)) {
    stop("dsHPC is not available")
  }
  
  status <- dsHPC.status(job_id)
  
  if (status$status == "COMPLETED" || status$status == "CACHED") {
    return(list(
      status = status$status,
      message = "Analysis completed. Use myPackage.getAnalysisResult() to retrieve results."
    ))
  } else if (status$status == "FAILED") {
    job_info <- dsHPC:::get_job_info(getOption("dsHPC.config")$connection, job_id)
    return(list(
      status = "FAILED",
      message = paste("Analysis failed:", job_info$error_message)
    ))
  } else {
    return(list(
      status = status$status,
      message = paste("Analysis is", tolower(status$status))
    ))
  }
}

# Function to get analysis result
myPackage.getAnalysisResult <- function(job_id) {
  if (!requireNamespace("dsHPC", quietly = TRUE)) {
    stop("dsHPC is not available")
  }
  
  # Check status first
  status <- dsHPC.status(job_id)
  
  if (!(status$status %in% c("COMPLETED", "CACHED"))) {
    stop(paste("Cannot retrieve result. Job status is", status$status))
  }
  
  # Get the result
  result <- dsHPC.result(job_id)
  return(result)
}
```

## Cross-Language Job Management

The same job management functions work identically for both R and Python jobs. This allows you to manage a mix of R and Python jobs using a single consistent interface.

```{r mixed_language_jobs}
# Submit both R and Python jobs
r_job <- dsHPC.submit(
  func = function(x) mean(x),
  args = list(x = rnorm(1000))
)

py_job <- dsHPC.submit_python(
  py_module = "numpy",
  py_function = "mean",
  args = list(a = rnorm(1000))
)

# Check status of both jobs
r_status <- dsHPC.status(r_job$job_id)
py_status <- dsHPC.status(py_job$job_id)

# Retrieve results
r_result <- dsHPC.result(r_job$job_id, wait_for_completion = TRUE)
py_result <- dsHPC.result(py_job$job_id, wait_for_completion = TRUE)

# List all jobs of any language
all_jobs <- dsHPC.list_jobs()
```

## Best Practices for Job Management

1. **Always check job status before retrieving results**
   ```r
   status <- dsHPC.status(job_id)
   if (status$status == "COMPLETED") {
     result <- dsHPC.result(job_id)
   }
   ```

2. **Use timeouts when waiting for job completion**
   ```r
   result <- dsHPC.result(job_id, wait_for_completion = TRUE, timeout = 3600)  # 1 hour
   ```

3. **Implement error handling for job failures**
   ```r
   status <- dsHPC.status(job_id)
   if (status$status == "FAILED") {
     # Handle the failure, perhaps by resubmitting or using a fallback method
   }
   ```

4. **Regularly clean up old jobs and results**
   ```r
   # Remove results older than 14 days
   dsHPC.clean_cache(days_to_keep = 14)
   ```

5. **Maintain job IDs for long-running processes**
   Store job IDs in a persistent location (e.g., a database or file) for monitoring long-running jobs that may span multiple R sessions.

## Conclusion

Effective job management is essential for making the most of the high-performance computing capabilities in dsHPC. The functions and patterns demonstrated in this vignette allow you to submit, monitor, and retrieve results from both R and Python jobs, providing a unified interface for computational tasks across languages. 