---
title: "Python Integration with dsHPC"
author: "David Sarrat González, Xavier Escribà-Montagut, Juan R. González"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Python Integration with dsHPC}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval = FALSE
)
```

## Introduction

This vignette demonstrates how to use the Python integration capabilities of the `dsHPC` package. The integration allows you to seamlessly submit Python functions as jobs through the DataSHIELD infrastructure, leveraging the power of the `reticulate` package for R-Python interoperability.

## Prerequisites

Before using the Python integration features, ensure you have:

1. A working Python installation
2. The `reticulate` R package installed
3. Any Python modules your code requires

```{r setup}
library(dsHPC)
library(reticulate)
```

## Initializing dsHPC

Before submitting any jobs, you need to initialize the dsHPC system:

```{r init}
dsHPC.init()
```

## Basic Python Integration

### Submitting a Simple Python Job

The `dsHPC.submit_python()` function allows you to submit a Python function to be executed as a job. Here's a simple example using NumPy:

```{r simple_job}
# Submit a job to calculate the mean of some numbers using NumPy
job <- dsHPC.submit_python(
  py_module = "numpy",
  py_function = "mean",
  args = list(a = c(1, 2, 3, 4, 5))
)

# Check the job status
status <- dsHPC.status(job$job_id)
print(status)

# Get the result when the job is complete
if (status$status == "COMPLETED") {
  result <- dsHPC.result(job$job_id)
  print(result)
}
```

### Specifying Python Environment

You can specify which Python environment to use:

```{r python_env}
# Using a specific Python executable
job <- dsHPC.submit_python(
  py_module = "numpy",
  py_function = "mean",
  args = list(a = c(1, 2, 3, 4, 5)),
  python_path = "/path/to/python"
)

# Using a virtual environment
job <- dsHPC.submit_python(
  py_module = "numpy",
  py_function = "mean",
  args = list(a = c(1, 2, 3, 4, 5)),
  virtualenv = "my_venv"
)

# Using a conda environment
job <- dsHPC.submit_python(
  py_module = "numpy",
  py_function = "mean",
  args = list(a = c(1, 2, 3, 4, 5)),
  condaenv = "my_conda_env"
)
```

## Working with Complex Data Types

### Passing Data Frames to Python

Data frames can be passed to Python as dictionaries:

```{r dataframe}
# Create a sample data frame
df <- data.frame(
  x = c(1, 2, 3, 4, 5),
  y = c("a", "b", "c", "d", "e"),
  stringsAsFactors = FALSE
)

# Submit job to process the data frame
job <- dsHPC.submit_python(
  py_module = "pandas",
  py_function = "DataFrame",
  args = list(data = df)
)
```

### Handling Complex Return Values

Python functions can return complex objects that are automatically converted to R objects:

```{r complex_returns}
# Function that returns a dictionary with various types
job <- dsHPC.submit_python(
  py_module = "os",
  py_function = "environ",
  args = list()
)

# Get the result (will be an R list)
result <- dsHPC.result(job$job_id)
```

## Custom Python Modules

You can also call functions from your own Python modules:

```{r custom_module}
# Assuming you have a custom module named "my_module.py"
# with a function called "process_data"
job <- dsHPC.submit_python(
  py_module = "my_module",
  py_function = "process_data",
  args = list(
    data = c(1, 2, 3, 4, 5),
    options = list(normalize = TRUE)
  )
)
```

## High-Performance Computing Options

### Setting Slurm Options

You can specify Slurm options for your Python jobs:

```{r slurm_options}
# Submit a job with specific Slurm requirements
job <- dsHPC.submit_python(
  py_module = "numpy",
  py_function = "linalg.eig",  # Computationally intensive function
  args = list(a = matrix(rnorm(100*100), 100, 100)),
  slurm_opts = list(
    partition = "normal",
    memory = "4g",
    time = "00:30:00",
    cpus = 4
  )
)
```

### Caching Results

By default, `dsHPC` caches results to avoid redundant computations. You can control this behavior:

```{r caching}
# Disable caching for this job
job <- dsHPC.submit_python(
  py_module = "numpy",
  py_function = "mean",
  args = list(a = c(1, 2, 3, 4, 5)),
  reuse_cached = FALSE
)
```

## Loading Required Modules

If your Python function requires additional modules, you can specify them:

```{r required_modules}
# Submit a job that requires multiple modules
job <- dsHPC.submit_python(
  py_module = "matplotlib.pyplot",
  py_function = "plot",
  args = list(x = c(1, 2, 3, 4, 5), y = c(10, 20, 15, 30, 25)),
  required_modules = c("numpy", "pandas")
)
```

## Conclusion

The Python integration in `dsHPC` provides a flexible and powerful way to execute Python code in a high-performance computing environment through DataSHIELD. By combining the strengths of both R and Python, you can create more versatile and efficient data analysis pipelines. 