---
title: "Python Integration with dsHPC"
author: "David Sarrat González, Xavier Escribà-Montagut, Juan R. González"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Python Integration with dsHPC}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval = FALSE
)
```

## Introduction

The `dsHPC` package provides seamless integration with Python through the `reticulate` package. This allows you to run Python code on HPC resources, with automatic conversion between R and Python data types.

## Prerequisites

Before using the Python integration features, ensure you have:

1. A working Python installation
2. The `reticulate` R package installed
3. Any Python modules your code requires

```{r setup}
library(dsHPC)
library(reticulate)
```

## Initializing dsHPC

Before submitting any jobs, you need to initialize the dsHPC system:

```{r init}
dsHPC.init()
```

## Basic Python Job Submission

The `dsHPC.submit_python()` function allows you to submit a Python function to be executed as a job. Here's a simple example using NumPy:

```{r simple_job}
# Submit a job to calculate the mean of some numbers using NumPy
job <- dsHPC.submit_python(
  py_module = "numpy",
  py_function = "mean",
  args = list(a = c(1, 2, 3, 4, 5))
)

# Check the job status
status <- dsHPC.status(job$job_id)
print(status)

# Get the result when the job is complete
if (status$status == "COMPLETED") {
  result <- dsHPC.result(job$job_id)
  print(result)
}
```

## Python Environment Configuration

### Using a Specific Python Executable

You can specify which Python interpreter to use:

```{r python_env}
# Using a specific Python executable
job <- dsHPC.submit_python(
  py_module = "numpy",
  py_function = "mean",
  args = list(a = c(1, 2, 3, 4, 5)),
  python_path = "/path/to/python"
)
```

### Using a Virtual Environment

```{r python_env}
# Using a virtual environment
job <- dsHPC.submit_python(
  py_module = "numpy",
  py_function = "mean",
  args = list(a = c(1, 2, 3, 4, 5)),
  virtualenv = "my_venv"
)
```

### Using a Conda Environment

```{r python_env}
# Using a conda environment
job <- dsHPC.submit_python(
  py_module = "numpy",
  py_function = "mean",
  args = list(a = c(1, 2, 3, 4, 5)),
  condaenv = "my_conda_env"
)
```

## Passing Data Between R and Python

The `dsHPC` package automatically converts R objects to Python and back. Here are some examples:

### Working with Vectors and Arrays

R vectors are converted to NumPy arrays:

```{r dataframe}
# R vector to NumPy array and back
r_vector <- c(1, 2, 3, 4, 5)
job <- dsHPC.submit_python(
  py_module = "numpy",
  py_function = "mean",
  args = list(a = r_vector)
)

# Result is an R scalar (3)
result <- dsHPC.result(job$job_id)
```

### Working with Data Frames

R data frames are converted to pandas DataFrames:

```{r dataframe}
# Create a sample data frame
df <- data.frame(
  x = c(1, 2, 3, 4, 5),
  y = c("a", "b", "c", "d", "e"),
  stringsAsFactors = FALSE
)

# Submit job to process the data frame
job <- dsHPC.submit_python(
  py_module = "pandas",
  py_function = "DataFrame",
  args = list(data = df)
)
```

## Python Function Caching

Just like with R functions, `dsHPC` includes an intelligent caching system for Python functions. This system identifies Python jobs based on:

1. The Python module name
2. The Python function name
3. The function's implementation (when available)
4. The exact arguments passed to the function

When the same Python function is called with the same arguments, the cached result is returned immediately without re-executing the function:

```{r caching}
# First execution of a Python function
job1 <- dsHPC.submit_python(
  py_module = "numpy",
  py_function = "mean",
  args = list(a = c(1, 2, 3, 4, 5))
)

# Status is COMPLETED, result is computed
job1$status
#> [1] "COMPLETED"

# Second execution with identical inputs uses the cache
job2 <- dsHPC.submit_python(
  py_module = "numpy",
  py_function = "mean",
  args = list(a = c(1, 2, 3, 4, 5))
)

# Status is CACHED, result is retrieved from cache
job2$status
#> [1] "CACHED"
```

### Implementation Details

For Python functions, dsHPC attempts to capture the function's bytecode for more accurate comparison. This means that functionally identical Python functions (even if they're imported differently or accessed through different paths) can be recognized as the same for caching purposes.

### Disabling Caching

If you need to run a Python function without using the cache (e.g., for time-sensitive data or non-deterministic functions), you can disable caching:

```{r caching}
# Disable caching for this job
job <- dsHPC.submit_python(
  py_module = "numpy",
  py_function = "mean",
  args = list(a = c(1, 2, 3, 4, 5)),
  use_cache = FALSE
)
```

## Working with Python Libraries

### Required Modules

You can specify additional Python modules that need to be imported:

```{r required_modules}
# Submit a job that requires multiple modules
job <- dsHPC.submit_python(
  py_module = "matplotlib.pyplot",
  py_function = "plot",
  args = list(x = c(1, 2, 3, 4, 5), y = c(10, 20, 15, 30, 25)),
  required_modules = c("numpy", "pandas")
)
```

## High-Performance Computing Options

### Setting Slurm Options

You can specify Slurm options for your Python jobs:

```{r slurm_options}
# Submit a job with specific Slurm requirements
job <- dsHPC.submit_python(
  py_module = "numpy",
  py_function = "linalg.eig",  # Computationally intensive function
  args = list(a = matrix(rnorm(100*100), 100, 100)),
  slurm_opts = list(
    partition = "normal",
    memory = "4g",
    time = "00:30:00",
    cpus = 4
  )
)
```

## Custom Python Modules

You can also call functions from your own Python modules:

```{r custom_module}
# Assuming you have a custom module named "my_module.py"
# with a function called "process_data"
job <- dsHPC.submit_python(
  py_module = "my_module",
  py_function = "process_data",
  args = list(
    data = c(1, 2, 3, 4, 5),
    options = list(normalize = TRUE)
  )
)
```

## Conclusion

The Python integration in `dsHPC` provides a flexible and powerful way to execute Python code in a high-performance computing environment through DataSHIELD. By combining the strengths of both R and Python, you can create more versatile and efficient data analysis pipelines. 